### 2019-07-31 paper/abstract.tex
updating draft and adding figures

The rotational evolution of FGKM dwarfs is still mostly unconstrained after
around 2-3 billion years.
This limit is set by the oldest open clusters whose members have precise light
curves, and therefore photometric rotation periods.
However, recent evidence suggests that the relationship between rotation
period and color/temperature/mass does not remain constant over large
timescales, contradicting the long-held assumption.
New results show that old clusters appear to have a flatter relationship
between rotation period and color than young clusters.
In this work we use the velocity dispersion of groups of stars as a proxy for
age in order to demonstrate that the relationship between rotation period and
effective temperature not only changes over time.
Not only this, the slope of the period-\teff\ relation even appears to invert:
at young ages G stars rotate more rapidly than K and M dwarfs, but at old ages
they rotate more rapidly.
In addition, we show that stars rotating just below the rotation period gap
are dynamically young, ruling out the possibility that the rotation period gap
is caused by incorrect period measurements or binary companions.

>>>>

% Word limit: 250

Due to a lack of precise ages for old stars with measured rotation periods,
the rotational evolution of FGKM dwarfs is still mostly unconstrained after
around 2-3 billion years.
% This limit is set by the oldest open clusters whose members have precise light
% curves, and therefore photometric rotation periods.
Recent evidence suggests that the relationship between rotation period and
color, temperature, and mass does not remain constant over large timescales,
contradicting the long-held assumption that underlies many gyrochronology
relations.
% For example, old clusters appear to have a flatter relationship between
% rotation period and color than young clusters.
In this work we use the kinematic ages of star populations to reveal the
evolving relationship between rotation period and effective temperature.
Not only does the period-color relation change over time, its slope appears to
{\it invert}: at young ages early K stars rotate more rapidly than late K
stars, but at old ages they rotate more rapidly.
% Although the gyrochronology relations fail to correctly predict the ages of
% old cool stars, we show that gyrochronology accurately predicts the ages of
% early K dwarfs, and young K dwarfs of all temperatures and those ages are
% consistent with ages predicted from age-velocity dispersion relations.
We find that the ages of field stars with measured rotation periods do not
exceed 4-6 Gyrs, resolving a long-standing degeneracy between age and magnetic
braking efficiency.
% It was previously unknown whether the upper-limit of the stellar rotation
% period distribution was created by a ceasura of magnetic braking (\ie old
% stars do not rotate more slowly after a critical age) or an inability to
% measure rotation periods for old stars.
% We show that the lack of long rotation periods is caused by enhanced
% measurement difficulty.
We also show that stars rotating just below the rotation period gap are
dynamically young, ruling out the possibility that the rotation period gap is
caused by incorrect period measurements or binary companions.
% Finally, we suggest that this gap may mark a critical point in the rotation
% period evolution of K dwarfs.
The main result of this paper could be produced by either an evolving
period-color relationship {\it or} mass-dependent dynamical heating.
Although we provide some evidence to suggest that the former effect is
dominant, we cannot be sure and therefore caution that the results presented
here are subject to this caveat.

### 2019-07-31 paper/intro.tex
updating draft and adding figures

\subsection{Gyrochronology}
The rotation periods of FGKM dwarfs increase with the square-root of time, and
this phenomenon was first observed several decades ago \citep{skumanich1972}.
This characteristic of main-sequence stars allows them to be dated via their
rotation periods, a practice known as gyrochronology.
This is convenient, since the ages of main-sequence stars are extremely
difficult, and at times impossible, to measure via the traditional age-dating
method of isochrone placement.
It has also been known for a long time that stars of the same age but
different masses have different rotation periods \racomment{(citation)}.
However, an underlying assumption behind many of the empirical gyrochronology
relations \citep[\eg][]{barnes2003, barnes2007, mamajek2008, meibom2011,
angus2015, angus2019} is that the relationships between rotation period and
color, and rotation period and age are {\it separable}.
In other words, the period-color relation is the same at all ages and the
period-age relation is the same at all colors.
It was recently shown that old stars appear to be too rapidly rotating for
their age, \citep{angus2015, vansaders2016, vansaders2018} and that a
mass-dependent modification to the classical \citet{skumanich1972} spin-down
law was needed in order to fit the data \citep{vansaders2016, vansaders2018}.
In addition, an even more recent analysis of middle-aged open clusters provide
further hints that the spin-down rate is mass-dependent \citep{curtis2019}.
Unfortunately, without rotation periods for members of old open clusters, or
any large group of old stars with precisely measured ages, it is extremely
difficult to calibrate the rotation-color relations at old ages.
For this reason, we turned to a population-based stellar age proxy: velocity
dispersion in order to investigate the period-temperature relations at old
ages.

@@

Stars are thought to be born in the thin disk of the Milky Way, orbiting the
center of the galaxy with a low out-of-plane, or vertical, velocity ($W$, or
$v_z$).
Young stars have relatively small vertical velocities, but gain angular
momentum in the vertical direction over time.
Although the cause of orbital heating is not well understood, interactions
with giant molecular clouds are thought to play an important role.

Stellar velocities have a long history of being used as an age proxy, with
several notable examples within stellar astronomy \citep[\eg][]{faherty2009,
west2011}.

Vertical {\it actions} are better age indicators than velocities, because
actions are calculated by integrating angular momentum over the Milky Way's
potential, and are therefore position invariant -- \eg\ a star will have the
same action at periapsis and apoapsis.
In contrast, a star's velocity will be different at periapsis and apoapsis --
so in this sense, {\it actions} are the natural quantities to use as age
proxies.
However, actions can only be calculated will full 6-dimensional position and
velocity information, yet most stars with measured rotation periods do not
have radial velocity measurements because they are \kepler\ targets and most
\kepler\ targets are relatively faint (between $\sim$11th and $~$18th
magnitudes).
For this reason, we used an alternative age proxy: velocity in the direction
of galactic latitude, \vb.
Since these stars only have proper motions, parallaxes and positions, with no
radial velocites, we could not calculate full 3D velocities or actions.
However, since the \kepler\ field is at low galactic latitude, \vb is a close
approximation to $v_z$.

Although the velocity of any individual star will only provide a weak age
constraint, the velocity dispersion of a group of stars can indicate whether,
on average, that group is old or young relative to other groups.
In this work we compare the velocity dispersions of groups of stars to
ascertain which groups are older and which younger and draw conclusions based
on the implied relative ages of populations.

>>>>

\subsection{Gyrochronology}
It has long been known that the rotation periods of FGKM dwarfs increase over
time \citep{skumanich1972}.
This characteristic of main-sequence stars allows them to be dated via their
rotation periods in a practice known as gyrochronology, which is convenient
since the ages of main-sequence stars are extremely difficult to measure via
the traditional age-dating method of isochrone placement.
It is also well established that stars of the same age but different {\it
masses} have different rotation periods \racomment{(citation)}, assumed to be
caused by the deeper convective zones, and therefore stronger magnetic dynamos
(and more efficient magnetic braking) in lower-mass stars
\footnote{Color is the directly observable quantity to which empirical
gyrochronology relation are usually calibrated to, rather than mass or
effective temperature.}.
However, an underlying assumption behind many of the empirical gyrochronology
relations \citep[\eg][]{barnes2003, barnes2007, mamajek2008, meibom2011,
angus2015, angus2019} is that the relationships between rotation period and
color, and rotation period and age are {\it separable}.
In other words, the period-color relation is the same at all ages and the
period-age relation is the same at all colors.
It was recently shown that old stars rotate more rapidly than a simple,
separable gyrochronology relation would predict \citep{angus2015,
vansaders2016, vansaders2018, metcalfe2019}, and that a mass-dependent
modification to the classical \citet{skumanich1972} spin-down law is required
to reproduce the data \citep{vansaders2016, vansaders2018}.
An even more recent analysis of middle-aged open clusters provides further
indication that the exponent of the period-age relation is mass-dependent.
New rotation period measurements of the 1.1 Gyr open cluster, NGC 6811,
revealed a flattened relationship between rotation period and color
\citep{curtis2019}.
In this cluster, the G dwarfs rotate at the same rate as the K dwarfs.

Unfortunately, the open clusters with rotation period measurements are mostly
young -- currently the oldest is 2.5 Gyr (NGC 6819).
Without rotation periods for precisely dated old stars, it is extremely
difficult to calibrate the relationship between rotation period and color at
old ages.
For this reason, we turned to a population-based stellar age indicator,
velocity dispersion, to investigate the period-temperature relations at old
ages.
% Since a separable period-age and period-color relation provided an adequate
% fit for the young stars, it was naturally assumed to apply to old stars too.

@@

Stars are thought to be born in the thin disk of the Milky Way, orbiting the
center of the galaxy with a low out-of-plane, or vertical, velocity ($W$, or
$v_z$).
Young stars have relatively small vertical velocities, but gain momentum in
the vertical direction over time.
Although the cause of orbital heating is not well understood, interactions
with giant molecular clouds are thought to play an important role
\racomment{(citations)}.
Although the velocity of any individual star will only provide a weak age
constraint, the velocity dispersion of a group of stars can indicate whether,
on average, that group is old or young relative to other groups.
In this work we compare the velocity dispersions of groups of stars to
ascertain which groups are older and which younger and draw conclusions based
on the implied relative ages of populations.
% Stellar velocities have a history of being used as an age proxy, with several
% notable examples within stellar astronomy \citep[\eg][]{faherty2009,
% west2011}.
Since the age-velocity dispersion relations (AVRs) are themselves still
actively being calibrated, it is difficult to directly compare gyrochronal
ages with kinematic ones.
However, regardless of the exact relation between velocity dispersion and
stellar age, it is expected to be monotonic relationship, therefore velocity
dispersion can be used effectively to {\it rank} groups of stars by age.
\racomment{Say some words here to prime the reader for the upcoming caveats.}

% Vertical {\it actions} are better age indicators than velocities, because
% actions are calculated by integrating angular momentum over the Milky Way's
% potential, and are therefore position invariant -- \eg\ a star will have the
% same action at periapsis and apoapsis.
% In contrast, orbital velocities are different at periapsis and apoapsis -- so
% in this sense, {\it actions} are the natural quantities to use as age proxies.
Vertical action is a better age indicator than vertical velocity, however both
vertical action and vertical velocity (\vz/W) can only be calculated with full
6-dimensional position and velocity information, yet most stars with measured
rotation periods do not have radial velocity measurements because they are
\kepler\ targets which are relatively faint (between $\sim$11th and $~$18th
magnitudes).
For this reason, we used an alternative age proxy: velocity in the direction
of galactic latitude, \vb.
The \kepler\ field is at low galactic latitude (b=5-20\degrees), so \vb is a
close approximation to \vz.
% Since the stars in our sample only have proper motions, parallaxes and
% positions, with no radial velocites, we could not calculate full 3D velocities
% or actions.

% \subsection{The degeneracy between gyrochronology and mass-dependent heating}

In this paper we tested the \citet{angus2019} gyrochronology relation which
was calibrated using the period-color relation of Praesepe and the period-age
relation of Praesepe and the Sun and is calibrated in \gaia\ \gcolor\ color.
% This relation was calibrated by fitting a 5th-order polynomial to the relation
% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
% and the Sun.
Although only calibrated using Praesepe and the Sun, this gyrochronology
relation was tested on NGC 6819, a 2.5 Gyr cluster, for which it predicted
accurate ages
In addition, the period-color relation of Praesepe is nearly identical to the
period-color relation of the Hyades, a cluster of around the same age: $\sim$
650 years \citep{douglas2019}.
However, this relation does {\it not} provide a good fit to NGC 6811, a 1.1
Gyr cluster \citep{curtis2019}.
The G dwarfs in NGC 6811 rotate at the same rate as the K dwarfs and it
appears as though the K stars have `stalled' -- their spin-down has been
halted.
The \citet{angus2019} gyrochronology relation assumes that the rotation
period-color relation of Praesepe is applicable to stars of all ages: the same
polynomial relation fit to Praesepe is used to describe the period-color
relation for all stars.
However, the NGC 6811 cluster suggests that this assumption does not hold past
around 1 Gyr.
If the period-color relation of the \citet{angus2019} gyrochronology model
{\it were} a perfect model for the rotational evolution of stars, then groups
of stars selected to be similar ages using this relation should fall on an
isochrone on both a CMD, and in period-\teff\ space.
Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
variations in metallicity and extinction, blur out the main sequence enough
that differences between stars in different age and rotation bins are not
easily discernable on the CMD, although there is still a general age and
rotation period gradient on the CMD, as seen in figures \ref{fig:CMD_cuts} and
\ref{fig:age_gradient}.
For this reason, we chose to use {\it kinematic} `isochrones', instead of
magnitude and color-based isochrones: although kinematics as an age indicator
is not necessarily as well calibrated to an absolute age scale as CMD
position, it can be very sensitive to {\it differences} in the ages of stellar
populations.
\begin{figure}
  \caption{
Main sequence stars with \mct\ rotation periods on the \gaia\ CMD with visual
    binaries removed.
    Points are colored by their gyrochronal age, according to the
    \citet{angus2019} gyrochronology relation.
    A general age gradient is visible across the main sequence.
}
  \centering
    \includegraphics[width=1\textwidth]{age_gradient}
\label{fig:age_gradient}
\end{figure}
Isochrones and stellar evolution tracks are highly dependent on choices made
about input physics and assumptions and have often been calibrated using
different types of stars.
As a result, different sets of models can have very different shapes on the
CMD, particularly at low masses.
Instead of relying on CMD position to age-date groups of stars, we opted to
explore age trends via kinematics.
Kinematic age-dating has the advantage of being relatively model independent,
or at least, having a very simple model: that velocity dispersion increases
over time.
This means that it is relatively easy to rank groups of stars by age: older
groups have a larger velocity dispersion.
It is less easy to rank-order groups of stars by age on the CMD because
stellar evolution is highly mass and metallicity dependent, and because
different stellar evolution models predict different ages for the same CMD
position.
This does not mean that it is not possible to calibrate gyrochronology using
isochrones, however it is not the focus of this study.

### 2019-07-31 paper/method.tex
updating draft and adding figures

\subsection{Calculating velocites}

We crossmatched the \mct\ catalog of stellar rotation periods, measured from
\kepler\ light curves with the \gaia\ DR2 catalog.
We added 2MASS K-band photometry from the Kepler Input Catalog
\citep{brown2011} and SDSS g-band photometry \racomment{(citations)}.

\begin{figure}
  \caption{
A color magnitude diagram in SDSS g and 2MASS K bands showing the
\citet{mcquillan2014} sample, colored by their rotation periods.
We excluded photometric binaries and subgiants from our analysis by removing
stars above the two dashed lines.
The rotation periods of binaries and subgiants do not follow a Skumanich-like
braking law and were removed from our sample.
% since we rely on a simple scaling
% between rotation period and age to make the argument that the rotation period
% is not caused by incorrect period measurements.
}
  \centering
    \includegraphics[width=1\textwidth]{CMD_cuts}
\label{fig:CMD_cuts}
\end{figure}

In order to explore the age of this stellar population from a rotation-period
standpoint, it was first necessary to remove visual binaries and subgiants
from the sample.
The rotation-period evolution of these two types of stars is generally
different to that of single stars which more usually follow a Skumanich-like
spin-down law.
Tidal and magnetic interactions between the two components of a binary system
can influence the rotation periods of both stars, and the expanding envelopes
of subgiants drive rapid spin-down through conservation of angular momentum.
We removed visual binaries from the sample by fitting a 6th-order polynomial
to the main sequence and raising it by subtracting 0.15 magnitudes to
approximate the line between the single-star and binary main sequences.
This polynomial is shown in figure \ref{fig:CMD_cuts}.
\racomment{Why did we do this in g-K?}
In order to remove subgiants, we removed stars brighter than 6th magnitude in
the g-band from the sample.

We used the {\tt Pyia} {\it Python} package to calculate stellar velocities.
{\tt Pyia} has built-in functionality for calculating velocity samples from
the full \gaia\ uncertainty covariance matrix via Monte Carlo sampling.
It therefore not only incorporates uncertainties on the \gaia\ positions
parallaxes and proper motions, it also accounts for the covariance between
these properties.
We calculated \vb\ for each star in the \mct\ catalog that survived the visual
binary and subgiant cuts.

>>>>

\subsection{The data}

We crossmatched the \mct\ catalog of stellar rotation periods, measured from
\kepler\ light curves, with the \gaia\ DR2 catalog.
Reddening and extinction from dust was calculated for each star using the
Bayestar dust map implemented in the dustmaps {\it Python} package
\citep{green2018}.
We estimated effective temperatures from dereddened \Gaia\ \gcolor\ color,
using an 8th-order polynomial relation calibrated using .... stars
\racomment{ask Jason for details}.
\begin{equation}
    \mathrm{T_{eff}} = 8960 -4802C + 1931C^2 -2446C^3 + 2669C^4 - 1324C^5 +
    301C^6 - 26C^7,
% 8959.8112335205078, -4801.5566310882568, 1931.4756631851196,
%           -2445.9980716705322, 2669.0248055458069, -1324.0671020746231,
%           301.13205924630165, -25.923997443169355]
\end{equation}
where C is \gaia\ \gcolor.

\begin{figure}
  \caption{
A \gaia\ color magnitude diagram showing the \citet{mcquillan2014} sample with
    extinction-corrected magnitudes, colored by their rotation periods.
We excluded photometric binaries and subgiants from our analysis by removing
stars above the two dashed lines.
% The rotation periods of binaries and subgiants do not follow a Skumanich-like
% braking law.
% The rotation period gradient across the main sequence is visible by eye in
%     this figure: young, rapidly rotating stars are located below the old,
%     slowly rotating stars.
% since we rely on a simple scaling
% between rotation period and age to make the argument that the rotation period
% is not caused by incorrect period measurements.
}
  \centering
    \includegraphics[width=1\textwidth]{CMD_cuts}
\label{fig:CMD_cuts}
\end{figure}
To explore the age of this stellar population from a rotation-period
standpoint, it was first necessary to remove visual binaries and subgiants
from the sample.
The rotation-period evolution of these two types of stars is generally
different to that of single stars which more usually follow a Skumanich-like
spin-down law.
Tidal and magnetic interactions between the two components of a binary system
can influence the rotation periods of both stars, and the expanding envelopes
of subgiants drive rapid spin-down through conservation of angular momentum.
We removed visual binaries and subgiants from the sample by applying cuts to
the color-magnitude diagram (CMD), shown in figure \ref{fig:CMD_cuts}.
We fit a 6th-order polynomial to the main sequence and raised it by 0.22 dex,
to approximate the division between single stars and visual binaries.
We removed all stars above this line from the sample.
We removed subgiants by eliminating stars brighter than 6th magnitude in
\gaia\ G-band from the sample.

We removed stars with negative parallaxes and parallax signal-to-noise ratios
below 10 from the sample.
We also removed a small number of stars fainter than 16th magnitude from the
sample.
We used the {\tt Pyia} {\it Python} package to calculate stellar velocities.
{\tt Pyia} has built-in functionality for calculating velocity samples from
the full \gaia\ uncertainty covariance matrix via Monte Carlo sampling.
It therefore not only incorporates uncertainties on the \gaia\ positions
parallaxes and proper motions, it also accounts for the {\it covariance}
between these properties.
We removed stars with absolute \vb\ uncertainties greater than 1 \kms\ from
the sample.
